A fault tolerant multi-server system for MMOGs

### 2011-07-06 trunk/textos/phd-prospectus/prospectus.tex
prospectus almost done

The rest of this document is organized as follows: Section \ref{sec:model} gives an overview of the system model and describes briefly some of its protocols; Section \ref{sec:futurework} presents the research directions to be followed next and Section \ref{sec:achievements} lists some of the accomplishments achieved by this research so far.

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed model} \label{sec:model}
%%%%%%%%%%%%%%%%%%%%%%%%%

We consider here persistent state real-time games with virtual environments where each player controls an avatar by issuing commands to it. Through the avatar, the player interacts with the virtual world and with objects present in it, such as avatars from other players. The basic operation of the game network support system must be as follows: when a player connects to the server, it must send the current state of the player's avatar and of the surroundings of its current location; after that, this player keeps receiving periodic state updates for the objects which are relevant to it (usually based on whether those objects may be seen by the player \cite{bezerra2008a3,ahmed2008dai,minson2005aim}). When a player issues a command (e.g. move his avatar from one place to another, pick up an object, attack another player's avatar etc.), it is sent as a message to the server, which processes it and decides its outcome, probably changing the state of the game, which is then broadcast to all involved players.

The next sections will describe the services the service must provide to the players connected to it (Section \ref{sec:services}), the specific requirements for a fault-tolerant distributed MMOG server system, considering the crash-stop failure model, as well as the proposed distributed architecture (Section \ref{sec:distserver}) and an algorithm devised to synchronize the game state between the replicas (Section \ref{sec:algorithm}).

@@

As already mentioned, in order to provide the services required by the MMOG, this work proposes the use of a geographically distributed system composed of voluntary nodes working together to form a distributed MMOG server. This distributed server must deliver, at least, the following services:

\begin{itemize}
	\item \textbf{Simulation} of the game, that is, the receiving of actions sent by the players and then processing their corresponding outcomes, which may result in changes to the game state;
	\item \textbf{Broadcasting} a portion of the current game state to the different players connected to it;
	\item \textbf{Storage} of the game state, that is, the combination of the states of each object in the virtual environment.
\end{itemize}

It is important to note that in real-time games (as opposed to turn-based games), due to the possibly fast-paced interaction of the players, it is critical to provide \textbf{timeliness} for the delivery of state update messages to them, so that they have the illusion of a common environment. Besides, it is necessary to provide \textbf{consistency} for the game state between each pair of players interacting with one another, which means that the game state they perceive must be as similar as possible. Although it is very hard to guarantee that every player will share the exact same view, it is important for their commands to be processed as if there was a single, strongly consistent game state, which is the one stored at the server.

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Distributed server system} \label{sec:distserver}

The main purpose of the system is to reduce or, ideally, eliminate the necessity of a central server for the game. In this work, the approach proposed to achieve this is based on principles similar to those of volunteer computing, where a potentially large set of contributors make their computers available to perform tasks requested by some remote entity. With such a system, although each of these nodes is probably much less powerful than a traditional mmog server, when working together they can sum up the cpu power and network link necessary to host one of these games.

However, several questions arise when dealing with a potentially large scale geographically distributed server system composed of voluntary nodes. First, as the nodes are volunteer, there is no guarantee that they will be available whenever they are needed, neither there is any guarantee regarding the amount of resources that each of these nodes makes available for the system -- be it cpu power, network bandwidth or storage capability. Group communication may also be a problem, since primitives for that -- such as network level multicast -- are not widely available on the Internet.

@@

In Section \ref{sec:statepart}, we describe how the game state will be distributed among the different server nodes; in Section \ref{sec:consprot}, a preliminary consistency management protocol, based on state machine replication, is presented.

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{State partitioning} \label{sec:statepart}

In order to split a game server into several smaller units executed by many volunteer nodes, it is used the idea of dividing the virtual environment in regions, each of which being managed by a different server. There are many approaches proposed in the literature, such as using fixed size cells which are grouped into regions, or using binary space partition trees, and so on. For the sake of generality, the architecture proposed here  does not decide how exactly the partitioning of the virtual environment will be performed. To cope with the possibility of failures, each server has several replicas, forming a server group.

The state partitioning is performed based on what objects are located in each region. The state of a object is managed, then, by the server group assigned to the region where that object is located. This way, the locality of data is explored: objects closed to each other are more likely to interact and, as they are located on the same server group, their interaction should be faster. Besides, by exploring the locality of the game data (state of the objects), the inter-group communication overhead is reduced.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Consistency protocol} \label{sec:consprot}

One of the assumptions here is that the commands from the players lead to a deterministic resulting game state. In other words, if the game starts in the state $S$, there is only one possible state $S'$ to be reached after a sequence $C=\{c_{1}, c_{2}, ..., c_{n}\}$ of commands. So, the game state may be modeled as a state machine, which is replicated across the different server nodes, as proposed in \cite{lamport1978ird,schneider1990ift,lampson1996hbh}. Therefore, a way to implement the distributed server system would be by creating such state machine and forward every command from the players to every server node in the same (total) order, so that they would always be able to calculate the correct game state\footnote{It might be the case that some servers could still be processing an older command, while others might have already processed newer inputs. Although this might lead to an old state, it would never lead to an incorrect state. To check which one is the newest, a version number considering, for instance, the number of commands processed -- since the commands arrive in the same order at the different servers -- would be enough.}, resulting in a fully replicated server system.

The problem with a fully replicated state machine approach for a distributed game server is that it would hardly scale well. As the number of servers increased, so would the cost of processing a command, as it would have to be totally ordered among other commands received by any of the other servers. Not only the number of control messages would grow and potentially saturate the servers' network bandwidth, but also the time needed for such total order to be achieved would possibly increase beyond a tolerable value.

We decided, therefore, to use a partial state machine replication scheme, so that only the server group responsible for a given state will have to process the commands that alter such states. To make this possible, each command must specify a target set -- defined either by the player or by a server --, which consists of the objects whose states are needed or altered by the command. As each object is assigned to a different server group, the protocol will then be able to infer which servers must process such command.

Each command will be processed by the group which contains its targets, being ordered among other commands who also are relevant to that same group. When a command has targets from different groups, than it must be ordered in all of the groups involved, i.e. all the groups which manage one of that command's targets.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{\textbf{Optimistic multicast}}

>>>>

The rest of this document is organized as follows: Section \ref{sec:mmogserver} describes the basic features of an MMOG server; Section \ref{sec:distserver} introduces our proposed distributed server model and describes briefly some of its protocols; Section \ref{sec:futurework} presents the research directions to be followed next and Section \ref{sec:achievements} lists some of the accomplishments achieved by this research so far.

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{MMOG server overview} \label{sec:mmogserver}
%%%%%%%%%%%%%%%%%%%%%%%%%

We consider here persistent state real-time games with virtual environments where each player controls an avatar -- a virtual object that represents him -- by issuing commands to it. Through the avatar, the player interacts with the virtual world and with objects present in it, such as avatars from other players. The basic operation of the game network support system must be as follows: when a player connects to the server, it must send the current state of the player's avatar and of the surroundings of its current location; after that, this player keeps receiving periodic state updates for the objects which are relevant to it (usually based on whether those objects may be seen by the player \cite{bezerra2008a3,ahmed2008dai,minson2005aim}). When a player issues a command to the avatar (e.g. move avatar from one place to another, pick up an object, attack another player's avatar etc.), it is sent as a message to the server, which processes it and decides its outcome, probably changing the state of the game, which is then broadcast to all involved players.

@@

In order to give to the players the sense of a persistent virtual world through which they can wander and interact with each other, the MMOG server must provide, at least, the following services:

\begin{itemize}
	\item \textbf{Simulation} of the game, that is, the processing of commands sent by players to their avatars, calculating the corresponding outcomes, which may result in changes to the game state;
	\item \textbf{Broadcasting} a portion of the current game state to the different players connected to it, so that their view is consistent with the changes that might have happened;
	\item \textbf{Storage} of the game state, which is the combination of the states of all objects in the virtual environment.
\end{itemize}

It is important to note that in real-time games (as opposed to turn-based games), due to the possibly fast-paced interaction of the players, it is critical to provide \textbf{timeliness} for the delivery of state update messages to them, so that they have the illusion of a shared environment. For the same reason, it is necessary to provide \textbf{consistency} for the game state between each pair of players interacting with one another, which means that the game state they perceive must be as similar as possible. Although it is very hard to guarantee that every player will share the exact same view, it is important for their commands to be processed as if there was a single, strongly consistent game state, which is the one stored at the server.

%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Distributed server system} \label{sec:distserver}

The main purpose of the system is to reduce or, ideally, eliminate the necessity of a central server for the game. The approach we propose to achieve this is based on principles similar to those of volunteer computing, where a potentially large set of contributors make their computers available to perform tasks requested by some remote entity. With such a system, although each of these nodes is probably much less powerful than a traditional MMOG server, when working together they can sum up the cpu power and the network bandwidth necessary to host one of these games.

However, several questions arise when dealing with a potentially large, geographically distributed server system composed of voluntary nodes. First, as the nodes are volunteer, there is no guarantee that they will be available whenever they are needed, neither there is any guarantee regarding the amount of resources that each of these nodes makes available for the system -- be it cpu power, network bandwidth or storage capability. Group communication may also be a problem, since primitives for that -- such as network level multicast -- are not widely available on the Internet.

@@

In Section \ref{sec:statepart}, we describe how the game state will be distributed among the different server nodes and, in \mbox{Section \ref{sec:consprot}}, a preliminary consistency management protocol, based on state machine replication, is presented. We also introduce, in Section \ref{sec:optmcast}, our group communication protocol on top of which our consistency management is performed.

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{State partitioning} \label{sec:statepart}

In order to split a game server into several smaller units executed by many volunteer nodes, it is used the idea of dividing the virtual environment in regions, each of which being managed by a different server. Different approaches can be found in the literature, such as using fixed size cells and grouping them to form the regions \cite{devleeschauwer2005dma}, or using binary space partition trees \cite{bezerra2009fgl}. To tolerate failures, each server has several replicas, forming a server group.

The state partitioning is performed based on which objects are located in each region. The state of an object is managed, then, by the server group assigned to the region where that object is located. This way, the locality of data is explored: objects close to each other are more likely to interact and, as they are located on the same server group, their interaction should be faster. Besides, by exploring the locality of the game data (state of the objects), the inter-group communication overhead is reduced.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Consistency protocol} \label{sec:consprot}

As mentioned before, the players interact by issuing commands and having them processed by servers, which calculate the resulting game state and send it to the players. Every command can alter the state of the game, possibly based on a previous state. We assume, however, that the commands lead to a deterministic resulting game state. In other words, if the game starts in the state $S$, there is only one possible state $S'$ to be reached after a sequence $C=\{c_{1}, c_{2}, ..., c_{n}\}$ of commands.

The game state could then be modeled as a state machine, which would be replicated across the different server nodes, as proposed in \cite{lamport1978ird,schneider1990ift,lampson1996hbh}, and where each command would be responsible for a state transition. One way of implementing the distributed server system would be by creating such state machine and forwarding all commands to every server node, in the same (total) order. As the state transitions are deterministic, every server would be able to calculate the correct game state independently\footnote{It might be the case that some servers could still be processing an older command, while others might have already processed newer inputs. Although this might lead to an old state, it would never lead to an incorrect state. To check which one is the newest, it would be enough to have a version number considering, for instance, the number of commands processed, since the commands arrive in the same order at the different servers.}. We would, then, have a fully replicated server system.

The problem with a fully replicated state machine approach for a distributed game server is that it would hardly scale well. As the number of servers increased, so would the cost of processing a command, as it would have to be totally ordered among other commands received by any of the other servers. Not only the number of control messages would grow and potentially saturate the servers' network bandwidth, but also the time needed for the total order of the commands to be achieved would possibly increase beyond a tolerable value.

We decided, therefore, to use a partial state machine replication scheme, so that only the server group responsible for a given state will have to process the commands that alter such state. To make this possible, each command must specify a \emph{target set} -- defined either by the player or by a server --, which contains the objects whose states are needed or altered by the command. As every object is assigned to a different server group, the protocol will then be able to infer which servers must process each command.

Instead of forwarding each command to every server, now only the servers that manage some object in the target set of a command would process it. The problem with such partial state machine replication is that, in order to execute some command, a group might need the previous state of an object managed by some other group. To solve this, a possibilty would be to have the groups requesting the states they need from other groups. Another solution, probably more latency-efficient, would be every group proactively sending the new states of the objects managed by it to other groups which might need them.

Sending new states to other groups proactively based on the possibility that they might need such states, however, may lead to all groups sending to each other the state changes of every object, degenerating to a fully replicated approach. Nevertheless, we can explore the fact that, in games, objects interact only with other objects situated nearby: we can define the regions of the virtual environment in such a way that only objects either in the same or in neighboring regions would be able to interact -- two regions are neighbors if, and only if, they share a border. As commands are executed by avatars, which are simply objects controlled by players, the result will be that, for each pair of objects in the target set of any command, these two objects will be either located in the same or in neighbor regions. This way, only groups managing regions that share a border would need to forward commands to each other\footnote{Although unlikely, it is possible to have a command sequence $C = \{c_1,c_2,\ldots,c_n\}$ where (1) every $c_i$ depends on the state resulting from $c_{i-1}$, (2) not every $c_i$ is processed by the same group and (3) not all of these groups manage neighbor regions. Nevertheless, the servers can detect such dependencies and forward the object states to the servers which need them. In the worst case, every group but one -- there is no dependency cycle -- would be waiting for another group, so the time needed to process a command would be proportional to the number of groups.}.

%Each command will be processed by the group which contains its targets, being ordered among other commands who also are relevant to that same group. When a command has targets from different groups, than it must be ordered in all of the groups involved, i.e. all the groups which manage one of that command's targets.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Optimistic multicast} \label{sec:optmcast}

As the servers are organized in groups, and every command is sent to one or more of such groups, we can use a multicast primitive. As commands seen by different servers must be handled in the same order, we need total order. As the commands sent by each player should be executed in the same order they were issued in, the multicast should provide FIFO ordering.

However, we want to minimize the inter-group communication overhead, so that our distributed server would be more scalable. Non-genuine multicast primitives \cite{schiper2009gng} require that control messages keep being exchanged between groups, even when there is no application level message to be sent. On the other hand, although genuine multicast algorithms minimize the number of messages exchanged, they impose a greater latecy on their delivery \cite{schiper2008ica}.
